# If Statements

If statements are a type of control flow structure.  Control structures help us control how many times code is executed, and when it will be executed.  This is helpful because sometime we only want our program to run if certain conditions are met.  

## If 

In R we can also execute commands only if certain conditions are met using the `if` operator.  This feature in R checks a logical value (`<value>`) and if it is `TRUE` then a sequence of commands within `{}` will be ran.  If `<value>` is `FALSE`, then the commands inside of `{}` will not be ran. 


```{r, eval = FALSE}
if(<value>){
  # Commands 
}

```


Below we have an example. The code will only be executed if `x` is a positive number. 

```{r}
x = 3

if(x>0){
  type = "positive"
}
type 
```

In the example above we have a relational operator which returns a logical value. This logical value was equal to `TRUE` so the code was executed. 

## If Else 

We can pair an  `if` statement with an `else` value.  After the `else` object we can define another sequence of commands inside of `{}`.  The `else` value is paired with the immediate previous `if` statement.  If this `if` statement is `FALSE` then the `else` code will run.  If the `if` statement is `TRUE` then the `else` code will not be executed. 

```{r}
x = -3

if(class(x) == "numeric"){
  type = "number"
} else {
  type = "not a number"
}
type
```

## Else If 

Sometimes we will want to do a sequence of checks that are all related, and we will only want code to run if the previous `if` statements were `FALSE` and another criteria is `TRUE`.  We can use `else if` to implement these rules. 

```{r}
x = -3


if(x>0){
    type = "positive"
} else if (x <0) {
    type = "negative"
} else if(x == 0){
    type = "zero"
} else {
    type = "Error"
}

type 
```

The command for the `if` statement will only run if `x>0`, and the rest of the code will not be implemented. 

The first `else if` commands will only execute if the first `if` statement was `FALSE` and `x<0`. 

The second `else if` commands will only execute if the previous `else if` and `if` statements were `FALSE` and `x==0`.  

If the previous `if` statement and all previous `else if` statements are `FALSE` then the `else` code will be executed. 


Here is another example with an `if else` chain.
```{r}
Toyfun <- function(X,Y,Do){
  if(Do == "Add"){
    Z = X+Y

  }else if(Do =="Subtract"){
    Z = X-Y

  }else if(Do =="Multiply"){
    Z = X*Y

    
  }else if(Do =="Penguin"){
    Z = c("<('' )")
    
  } else{
    Z = c(X,Y)
  }
  
  return(Z)
}
Toyfun(2,4,"Add")
Toyfun(2,4,"Subtract")
Toyfun(2,4,"Penguin")
Toyfun(2,4, "Typo")
```


## Ifelse

We have a shortcut function that can help us keep our code more succinct using the `ifelse()` function.  The first argument of this function is a logical value.  the second and third arguments are what to do if the value is `TRUE` or `FALSE`, respectively.   

```{r}
x <- 3
type <- ifelse(x>0,  "positive", "nonpositive")
type 
```


We can also pass `ifelse()` a vector, and it will check the logical condition for each element of the vector. For example, in the `mtcars` data set, we can find the proportion of cars have `mpg > 25` and `hp > 60`

```{r}
fast_efficient <- ifelse(mtcars$mpg > 25 & mtcars$hp>60, TRUE, FALSE)
sum(fast_efficient)/length(fast_efficient)
```


## Nested If Chains

We can make `if-else` chains nested within each other. 

```{r}
x <- 105
if(x>0){
  if(x>100){
     type = "large positive number"
  } else {
     type = "positive number"
  }
} else if(x<0) {
  type = "negative number"
} else if(x==0){
  type = "zero"
}else {
  type = "Error"
}

type 
```

## Additional Resources {-}

- [Chapter 13 of "R for Programming in Data Science"](https://bookdown.org/rdpeng/rprogdatascience/functions.html#the-...-argument) 
- Videos:
    + [Control Flow 1 | if Statements](https://ucr.yuja.com/V/Video?v=2369005&node=8488277&a=1089164180&autoplay=1)
    + [Control Flow 2 | if else Statements](https://ucr.yuja.com/V/Video?v=2369004&node=8488276&a=68950775&autoplay=1)
    + [Control Flow 3 | else if Statements](https://ucr.yuja.com/V/Video?v=2369003&node=8488275&a=577245814&autoplay=1)
    + [Control Flow 4 | ifelse Function](https://ucr.yuja.com/V/Video?v=2369002&node=8488274&a=2043700498&autoplay=1)
    + [Control Flow 5 | switch Function](https://ucr.yuja.com/V/Video?v=2369001&node=8488273&a=523725749&autoplay=1)