# ggplot2


## Tidyverse and ggplot2

For this document you will need to install and load the package `ggplot2` which is part of `tidyverse`.  To install the package refer to Section \@ref(packages). 

```{r warning=FALSE, message=F}
library(tidyverse)
```


The tidyverse is a collection of packages that share a unique underlying philosophy, frame work, and syntax.  There are approximately 20 tidyverse packages, but the core ones are `ggplot2`, `dplyr`, `tidyr`, `readr`, `purr`, `tibble`, `stringr`, and `forcats`.  You can install these packages individually or all at once using by simply using the command `install.packages("tidyverse")`. 

We will be focusing on the `ggplot2` package for now. The functions and topics that we that we will be covering can be found in @wickham2009elegant. 


## Intro to ggplot2

The package `ggplot2` is used to produce graphics using a different system then traditional "Base R". It has an underlying grammar or style that is different that is based on the "grammar of graphics". 


### What is the grammar of graphics?{-}

The structure of the "grammar of graphics" aims to create graphics using different layers. The grammar tells us how to map aesthetic attributes (color, shape, size) of a geometric object (points, lines, bars).  We might also have a statistical transformation of the data and information about the coordinate system. Faceting is also considered. The combination of these elements make up a "graphic". 

All plots are composed of **data** and a **mapping**.

- data: information you want to visualize

- mapping: description of how the variables are mapped to aesthetic attributes. 


There are five mapping components: 

- **Layer**: a collect of geometric elements and statistical transformations. 

    - Geometric elements are called **geoms**, they are what you see (points, lines, etc.)
    
    - Statistical transformations are called **stats**, they summarize the data (counting observations, makeing models)
    

- **Scales**: a map of the values in the data space to the aesthetic space. 

     - Modifications of the color, shape, size
     
     - How to draw legends and axes
     
- **Coordinate** System: called **coord**, describe how data coordinates are mapped to the plane of the graphic. 

     - Usually use Cartesian coordinate system. 
  

- **Facet**: How to break up and display subsets of data as small multiples. 

- **Theme**: control other points of display, like font size, and background color. 


### ggplot2 versus other R Graphics {-}

- Base R: graphics work like a "pen on paper"

- Grid: started in 2000, provided a richer system of graphics tools that are more complex. 

- Lattice: a new system proposed (2008ish).  Uses grid graphics to implement a trellis system that populates more plotting details automatically

- ggplot2: started 2005, an attempt to combine the good things about lattice and base R together using the "grammar of graphics"


## Getting started: Kep Components

Every ggplot2 plot has 3 key components:

1) Data

2) A set of **aes**thetic mappings between variables in the data and visual properties

3) At least one layer which described how to render each observation.  Layers are usually created with a **geom** function. 


A Basic Example: 

```{r}
ggplot(mpg,                        # Data
       aes(x = displ, y = hwy))+   # aesthetic mapping
  geom_point()                     # geom layer 
```

## Adjust aesthetic attributes by variable 

We can adjust aesthetic attributes by variable using the `aes()` function and the `mapping` argument.  

```{r, eval = F}
# Adjust the color for each point by class
aes(x = displ, y = hwy, color = class)

# Adjust the shape for each point by drv
aes(x = displ, y = hwy, shape = drv)

# Adjust the size of each point by cyl 
aes(x = displ, y = hwy, shape = cyl)
```



```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + 
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, shape = factor(cyl))) + 
  geom_point()
```

## Adjust aesthetic attributes for the entire plot 

Notice that the plot changes based on if the `color` is assigned outside or inside of the `aes()` function.  This is because the `aes()` function is used to map the data to an aesthetic property of the map.  Inside of `aes()` we let the aesthetic argument (color, shape, etc) equal a variable in the data set.  The different values for that variable control what happens to that aesthetic.  If we assign aesthetic value outside of the `aes()` function then we are just assigning a value for the entire data set and do not use column names of the data. 

```{r}

# Change aesthetic attributes according to some variable
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = class))

# Attempt to change all points to blue
# Will not run correctly because colour = "blue" is inside aes() 
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(color = "blue"))

# Change all points to blue, works correctly
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(color = "blue")
```

## Faceting 

Faceting creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset. You can faceting using `facet_wrap()` which is more flexible and automatically determines how to organize the data.  Or, you can manually determine the facet structure using `facet_grid()`.

```{r}

# Facet Wrap:  Let ggplot decide how many rows and columns to make
# Only give one variable to break data up by ~ Variable
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  facet_wrap(~class)

# Facet Grid: Number of rows and columns by number of unique values
# for variable 1 and variable 2 
# Row variable ~ Column variable. 
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  facet_grid(drv~class)
```


## Some example plots

```{r}

# Example 1
ggplot(diamonds)+
  geom_boxplot(mapping=aes(y = price, x = cut, color = cut))

# Example 2
ggplot(diamonds)+
  geom_point(mapping=aes(y = price, x = depth, size = carat, 
                         color = cut), pch = 1)+
  labs(title = "A Title is On This Plot")

# Example 3
subgroups_diamonds <- diamonds %>%
  group_by(cut, clarity) %>%
  summarize(n = length(cut))

ggplot(data = subgroups_diamonds) +
  geom_tile(mapping = aes(x = cut, y = clarity, fill = n))+
  scale_fill_gradient(low="white", high="purple")


# Example 4: Method 1
subgroups_diamonds2 <- diamonds %>%
  group_by(cut, color) %>%
  summarize(n = length(cut))

ggplot(data = subgroups_diamonds2)+
  geom_bar(mapping = aes(x = cut, y = n, fill = color),
           position="fill", stat = "identity")

# Example 4: Method 2
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill=color), 
           position = "fill")


# Example 5
plot_data <- diamonds %>%
  group_by(clarity) %>%
  summarize(m =mean(price))

ggplot(plot_data, aes(x = clarity, y = m)) +
  geom_bar(stat = "identity",
           show.legend = FALSE,
           color = "red",
           size = 1,
           alpha = .5)


# Example 6
ggplot(diamonds, aes(x = clarity, group = cut, fill = cut)) +
  geom_bar(position = "dodge")


# Example 7
  ## creating dataset #1
  data1 <- diamonds %>%
    group_by(clarity) %>%
    summarise(m = mean(price))

  ## creating dataset #2
  data2 <- diamonds %>%
    group_by(cut, clarity) %>%
    summarise(m = mean(price))

  ## graphing data points from 2 different datasets on one graph
  ggplot() +
    geom_point(data = data1, aes(x = clarity, y = m), color = "blue", pch = 8) +
    geom_point(data = data2, aes(x = clarity, y = m), pch = 1)

```


## Where can I learn more?

- https://ggplot2-book.org/index.html

- https://www.stat.auckland.ac.nz/~ihaka/downloads/Interface98.pdf

- https://hbctraining.github.io/Intro-to-R-flipped/cheatsheets/data-visualization-2.1.pdf
